Response to reviewers

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
revised sex and race paragraphs

\bigskip
\begin{abstract}
Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package called \texttt{yowie}.
\end{abstract}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ID: the respondents' ID.
\item
  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
\item
  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
\item
  BLACK: binary variable, 1 indicates black and 0 otherwise.
\item
  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
\item
  HGC: the highest grade completed.
\item
  UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).
\end{enumerate}

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, we use a single categorical race variable instead of the two binary race variables. We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to re-create the unemployment rate variable.

@@

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organized such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (742 to be specific). The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data shows:

@@

Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).

For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}. It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database. We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}. Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al. 2020). From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al. 2020). Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention. Similarly, this also relates to race, as reported in the database. When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix. In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.

@@

The respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all respondents were in this range in the data we extracted. Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the \texttt{demog\_nlsy79}, i.e., the data with the survey years 1979 sample. Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest the demographic data we had is consistent with the sample information in the database.

@@

\caption{\label{tab:gender-race-table}Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
\midrule
Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{id}: A unique individual's ID number.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{gender}: Gender of the individual, FEMALE and MALE.
\item
  \texttt{race} : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
\item
  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
\item
  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{gender}: Gender of the individual: FEMALE and MALE.
\item
  \texttt{race}: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
\item
  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\item
  \texttt{njobs}: Number of jobs that an individual has.
\item
  \texttt{hours:}: The total number of hours the individual usually works per week.
\item
  \texttt{stwork}: The year when the individual started to work.
\item
  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
\item
  \texttt{exp}: Work experience, i.e., the number of years of working.
\item
  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
\item
  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
\end{enumerate}

@@

Any longitudinal dataset used for education should have a sufficiently reproducible process to be updated with new data. The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately. This paper has illustrated the necessary steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research. In the first stage, we showed the steps performed to get the data from the NLSY79 database. The data format was converted to a tidy format for more flexibility in cleaning and exploring. Initial data analysis was conducted to investigate and screen the quality of the data. We found and provided a fix for many anomalous observations in wages using a robust linear regression model. The refreshed data is compared with the original set using a variety of numerical summaries and graphics. The current subset is made available in a new R package, called \texttt{yowie}.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and being black leads to lower wages. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

@@

\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
Grimshaw, Scott D. 2015. {``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14. \url{https://doi.org/10.1080/00031305.2015.1081106}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. {``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1). \url{https://doi.org/10.5070/T511103589}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

>>>>

\bigskip
\begin{abstract}
Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology. Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection. It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today. This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s. The data is useful for teaching modeling and exploratory analysis of longitudinal data. Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles. The NLSY79 database has been continuously updated through to 2018, so new records are available. Here we describe our journey to refresh the wages data, and document the process so that the data can be regularly updated into the future. Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated. We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching. Three new data sets and the code to produce them are provided in the open source R package called \texttt{yowie}.
\end{abstract}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ID: the respondents' ID.
\item
  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
\item
  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
\item
  BLACK: binary variable, 1 indicates Black and 0 otherwise.
\item
  HISPANIC: binary variable, 1 indicates Hispanic and 0 otherwise.
\item
  HGC: the highest grade completed.
\item
  UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).
\end{enumerate}

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, we use a single categorical race variable instead of the two binary race variables, for reasons detailed below. We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to refresh the unemployment rate variable.

@@

To get the data into tidy form (Wickham 2014), it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, \texttt{new\_data\_qnames}, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns. The raw data consequently has a large number of columns (742 to be specific). The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data show the format problems:

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).

For \texttt{sex} and \texttt{race}, we have simplified the original names provided in the raw data, \texttt{SAMPLE\_SEX\_1979} and \texttt{SAMPLE\_RACE\_78SCRN}, respectively. However, it is important to note that the use of ``sex'' is probably not correct. This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al. (2016). When using the dataset in the classroom, the educators might include discourse on the use of terms ``sex'' and ``gender''. Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification. From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources. Kennedy et al. (2020) provides a helpful discussion on these topics.

Similarly, there are issues for the \texttt{race} variable. The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category. in addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race. This level of detail is not available in the database. There is one variable with only three categories, and an individual can be a member of only one. Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.

Further, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on). We decide to include both

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix. In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.

@@

The respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all respondents were in this range in the data we extracted. Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their sex (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the \texttt{demog\_nlsy79}, i.e., the data with the survey years 1979 sample. Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest the demographic data we had is consistent with the sample information in the database.

@@

\caption{\label{tab:gender-race-table}Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Sex & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
F & 1,561 (25\%) & 1,002 (16\%) & 3,720 (59\%) & 6,283\\
M & 1,613 (25\%) & 1,000 (16\%) & 3,790 (59\%) & 6,403\\
\midrule
Total & 3,174 (25\%) & 2,002 (16\%) & 7,510 (59\%) & 12,686\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{id}: A unique individual's ID number.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race} : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
\item
  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
\item
  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race}: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{grade}: Integer value of the highest grade completed corresponding to the survey year.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\item
  \texttt{njobs}: Number of jobs that an individual has.
\item
  \texttt{hours:}: The total number of hours the individual usually works per week.
\item
  \texttt{stwork}: The year when the individual started to work.
\item
  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
\item
  \texttt{exp}: Work experience, i.e., the number of years of working.
\item
  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
\item
  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
\end{enumerate}

@@

Any longitudinal dataset used for education should have a sufficiently reproducible process to be refreshed with new data. The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately. This paper has illustrated the necessary steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research. In the first stage, we showed the steps performed to get the data from the NLSY79 database. The data format was converted to a tidy format for more flexibility in cleaning and exploring. Initial data analysis was conducted to investigate and screen the quality of the data. We found and provided a fix for many anomalous observations in wages using a robust linear regression model. The refreshed data is compared with the original set using a variety of numerical summaries and graphics. The current subset is made available in a new R package, called \texttt{yowie}.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that wages on average differ by race category. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

@@

\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
Grimshaw, Scott D. 2015. {``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14. https://doi.org/\url{https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\vadjust pre{\hypertarget{ref-SAGER}{}}%
Heidari, S., T. F. Babor, P. De Castro, S. Tort, and M. Curno. 2016. {``Sex and Gender Equity in Research: Rationale for the {SAGER} Guidelines and Recommended Use.''} \emph{Research Integrity and Peer Review} 1 (2). https://doi.org/\url{https://doi.org/10.1186/s41073-016-0007-6}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. {``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1). https://doi.org/\url{https://doi.org/10.5070/T511103589}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-doh}{}}%
Office of Management and Budget. 1997. {``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.''} \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. https://doi.org/\url{https://doi.org/10.18637/jss.v040.i01}.

### 2022-05-11 paper/paper.Rmd
revised sex and race paragraphs

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, we use a single categorical race variable instead of the two binary race variables. We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to refresh the unemployment rate variable.

@@

For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively. However, it is important to note that the use of "sex" is probably not correct. This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in  Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using]. From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using]. Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.

For the `race` variable, we acknowledge a current standard of how race should be reported in a dataset [@doh]. According to this standard, there should be at least five categories for race reported in a data, and one can be identified into more than one race category. Also, Hispanic or Latino is identified as an ethnic group rather than a race. In the refreshed data, we could not attempt to report the race and ethnicity with this standard as we did not have any additional resources to do so, and hence we only used the category as in provided in the database. This fact could be a suggestion for the Bureau of Labor of Statistics as the data provider to update the race reporting to today's standard. Similar to the "sex variable," the educators should also emphasize that when analyzing this variable, one should keep in mind that the purpose is to study racism rather than race.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and being Black leads to lower wages. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

>>>>

We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published). We also modify some variables. For example, we use a single categorical race variable instead of the two binary race variables, for reasons detailed below. We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked. We do not attempt to refresh the unemployment rate variable.

@@

For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively. However, it is important to note that the use of "sex" is probably not correct. This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in @SAGER. When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender". Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification. From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.  @kennedy2020using provides a helpful discussion on these topics. <!--Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. -->

Similarly, there are issues for the `race` variable. The current US Federal guidelines [@doh] state that there should be at least five categories for race, and an individual should be able to identify as more than one race category. in addition, "Hispanic" or "Latino" should be considered to be  an ethnic group rather than a race. This level of detail is not available in the database. There is one variable with only three categories, and an individual can be a member of only one. Thus, the `race` variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.

@@

The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns. More education leads to increasingly higher wages, which is a satisfying result for educators. It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that wages on average differ by race category. This is a disturbing finding because there is no rationale for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

### 2022-05-11 response/response2.Rmd
revised sex and race paragraphs

* We interpret this to mean we should change the variable name in the cleaned data to be `sex`. We have done this and then "sex" is uniformly used throughout the paper. Even though we have changed the `gender` variable to be `sex`, we keep the explanation of the importance for the educators to emphasize the discussion between the usage of `sex` and `gender` because in many other cases and datasets, the usage between `sex` and `gender` are often interchangeable. XXX DONE

@@

* It would be nice if the database contained a broader description of race, but it only has a single variable for the race. For this reason, our cleaned data reflects this and has a single variable for the race. We have explained why we used these categories and suggested the data provider adjust the race variable to use the current standard. XXX DONE

>>>>

* We have revised the language, made a note for educators on usage, and referred to the SAGER guidelines. XXX DONE

@@

* We have revised the language, referenced the Federal standard, and explained the reasoning for using a single variable.  XXX DONE
